Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2024-01-05 doc/JAIR/jair-paper.tex
Minor polishing

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
(det-DNNF) introduced by
\citeA{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
represent a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

To ensure soundness of the abstract CPOG proof system, as well as
correctness of its concrete implementation, we formally verified the
proof system as well as versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a  CPOG file gives strong
assurance that the proof and the model count are correct. Our
experience with developing a formally verified proof checker has shown
that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of
requirements that guarantee soundness.  In fact, our efforts to
formally verify our proof framework exposed subtle conditions that we had
to impose on our partitioned sum rule.

@@

This paper is an extended version of one published at the 2023
Conference on the Theory and Application of Boolean
Satisfiability~\cite{bryant:sat:2023}.  It provides greater detail of
the algorithms, the formal verification, and the experimental results.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{heule:cade:2013,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
both
that it
satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

@@

\section{Logical Foundations}
\label{section:logical:foundations}

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{prop}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation}\\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product}\\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum} 
\end{eqnarray}
\end{prop}
As is described in \ref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.

@@

We can now define and compare two related representations:
\begin{itemize}
\item A det-DNNF graph can be viewed a POG with negation applied only to variables.
\item A dec-DNNF graph is a det-DNNF graph with the further restriction that any
  sum node $\nodeu$ has exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\phi_{\nodeu_b})$ has $\assign(x)=b$, for $b \in \{0,1\}$.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating dec-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

Let $\exvarset(\theta)$
denote the set of extension variables occuring in formula $\theta$.
For any assignment $\assign$ to the variables in $\varset$, the defining
clauses induce a unique assignment $\eassign$ to the variables in $\varset
\cup \exvarset(\theta)$.  For Boolean formula $\phi$ over variables
$\varset$ and clausal
formula $\theta$ over the variables $\varset \cup \exvarset(\theta)$, we say that $\phi$ is {\em
  effectively equivalent} to $\theta$, written
$\phi \eequiv \theta$,
when for any assignment
$\assign$ to the variables in $\varset$, assignment $\assign$ is a model of $\phi$ if
and only if its extension $\eassign$ is a model of $\theta$.
Starting with  $\theta$ equal to input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \eequiv \theta$.

@@

Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
%% The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.

@@

\section{Generating CPOG from dec-DNNF}
\label{section:generating:cpog}

@@

\section{A Formally Verified Toolchain}
\label{section:formally-verified-toolchain}

@@

The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{section:logical:foundations}. Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas. \lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}. Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}. Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$. We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.

>>>>

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
(det-DNNF) introduced by
\citeA{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
represent a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean \textsc{and} or \textsc{or} operation.  Restrictions
are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting} framework~\cite{kimmig:jal:2017} describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

To ensure soundness of the abstract CPOG proof system, as well as
correctness of its concrete implementation, we formally verified the
proof system as well as versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a  CPOG file gives strong
assurance that the proof and the model count are correct. Our
experience with developing a formally verified proof checker has shown
that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of
requirements that guarantee soundness.  In fact, as described in Section~\ref{sect:lean:subtle-condition}, our efforts to
formally verify our proof framework exposed subtle conditions that we had
to impose on our partitioned sum rule.

@@

This paper is an extended version of one published at the 2023
Conference on the Theory and Application of Boolean
Satisfiability~\cite{bryant:sat:2023}.  It provides much greater detail about
the algorithms, the formal verification, and the experimental results.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{heule:cade:2013,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
In particular, our proofs construct a new propositional formula, and we must verify
both
that it
satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{sect:formally-verified-toolchain}.

@@

\section{Logical Foundations}
\label{sect:logical:foundations}

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{prop}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation}\\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product}\\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum} 
\end{eqnarray}
\end{prop}
As is described in \ref{sect:formally-verified-toolchain}, we have proved these three equations using \lean{}.

@@

We can now define and compare two related representations:
\begin{itemize}
\item A det-DNNF graph can be viewed a POG with negation applied only to variables.
\item A dec-DNNF graph is a det-DNNF graph with the further
  restriction that any sum node $\nodeu$ has exactly two children
  $\nodeu_1$ and $\nodeu_0$, and it has an associated \emph{decision variable} $x$.
  For $b \in \{0,1\}$, node $\nodeu_b$ can be a terminal node with variable $x$, where the polarity of the edge from $\nodeu$ to $\nodeu_b$ is
  negative for $b=0$ and positive for $b=1$.
  Alternatively, $\nodeu_b$ can be a product node having either literal $\obar{x}$ ($b=0$), or literal $x$ ($b=1$) as one of its arguments.
  Either form implies that any total assignment $\assign \in \modelset(\phi_{\nodeu_b})$
  has $\assign(x)=b$, for $b \in \{0,1\}$.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating dec-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

Let $\exvarset(\theta)$
denote the set of extension variables occuring in formula $\theta$.
For any total assignment $\assign$ to the variables in $\varset$, the defining
clauses induce a unique assignment $\eassign$ to the variables in $\varset
\cup \exvarset(\theta)$.  For Boolean formula $\phi$ over variables
$\varset$ and clausal
formula $\theta$ over the variables $\varset \cup \exvarset(\theta)$, we say that $\phi$ is {\em
  effectively equivalent} to $\theta$, written
$\phi \eequiv \theta$,
when for any assignment
$\assign$ to the variables in $\varset$, assignment $\assign$ is a model of $\phi$ if
and only if its extension $\eassign$ is a model of $\theta$.
Starting with  $\theta$ equal to input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \eequiv \theta$.

@@

Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
%% The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be defined in order,
with any clause identifier $C'$ given in the hint when adding clause $C$ having $C' < C$.

@@

\section{Generating CPOG from dec-DNNF}
\label{sect:generating:cpog}

@@

\section{A Formally Verified Toolchain}
\label{sect:formally-verified-toolchain}
\label{sect:lean:subtle-condition}

@@

The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{sect:logical:foundations}. Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas. \lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}. Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}. Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$. We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
